Case-based Arabic Morphological Analysis

### 2010-12-16 tex/sarf.aclhlt.tex
save from diff

\begin{document}
\maketitle
\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, and in particular in the context
of the Arabic language, annotations resulting from 
an ongoing branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case based morphological 
analyzer for Arabic text. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses compositional non-deterministic finite state machines 
to analyze a stream of text. 
Each alive machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The letter \RL{'a} may be prefix and the word means ``I praise him''.
The letter \RL{'a} may also be part of the stem 
\RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Another difficulty is that  Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a delimiter such as a space in between. 
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often in text, and greatly increase the
difficulty of tokenization.

@@

We find evidence to our hypothesis in~\cite{Maamouri:10} where 
a new corpus was added to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}. 
The addition required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new genre of data challenged the existing data processing and 
annotation guidelines. 
Alternative guidelines came to place where the high level NLP 
task interacted with the morphological analyzer. 
This led to refining SAMA~\cite{Kulick:10}. 
The work of Habash~\shortcite{Habash:06} reports that different 
types of morphological analyses behave differently on the same case 
study. 
For one, the analyzers may find different solutions. 
The analyzers may present the solutions in different orders. 
The presented solutions and the order of presentation may not be 
appropriate for the case study. 
We find strong evidence in our results 
that a case-based morphological analyzer that allows a case-based 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

@@

In this paper, we present Sarf, a {\em novel case-based efficient
morphological analyzer} that uses compositional 
non-deterministic Automata driven by a case based controller.
Each alive machine in Sarf represents a valid morphological analysis. 
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

Sarf uses a {\em recursive} concatenative analysis with a novel 
refinement. 
Sarf introduces recursive affixes in order to
betterr retain part of speech information and enhance the 
efficiency of affix matching.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each narration into
its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
connected toghether they form complex names of narrators. 
For example,
narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely as long 
as the text is valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure. 
We also showcase that the results of the accurate high level 
abstraction
results of the case-based analysis was not affected by a major
refinement to the analyzer that increased low level analysis 
accuracy.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of Automata generated by a compiler
from Xerox rules can not be controlled and also reports a 
difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix tansducers. 
With Sarf it is easy to control the FSA with a custom user case 
controller. 
Up to our understanding, Beesley  
requires recompiling the Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions 
that partition text into morphemes.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
The algorithmic changes in SAMA were 
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example construction of the affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
\transtrue

@@

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state. 
Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

@@

Morphological analysis partitions a given Arabic word
into a set of morphemes.
A morpheme is either an affix or a stem. 
We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of prefixes, suffixes and stems and assigns a category
to each morpheme. 
The relation between morphemes and categories is one 
to many.

A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of the string is a 
valid morphological analysis if  and only if we have
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant.
When users try to add their custom morphemes to these
lists they may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
is that of the second morpheme in the pair.

Formally we define an affix $f$ recursively to be either
an atomic affix $a$ or an atomic affix concatenated
to an affix and we write $f: a | a f$. 
Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Sarf only keeps atomic affixes and their categories.

\subsection{Affix linear transducer}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into a structure similar to those
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two structures are directed acyclic graphs. 
The absence of cycles reduces the 
traversal of the affix transducer to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix transducers encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the transducer corresponds to the last letter 
of a prefix in the prefix and suffix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsection{Stem linear transducer}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names. 
We obtained the set of proper names from online 
and biblical sources.

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is also linear where the accept
states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Non-deterministic composition of transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic composition 
of the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all linear and do not have cycles as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for brevity and pedagogical purposes.

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{An abstraction of the non-deterministic composition of 
    the prefix, stem and suffix transducers.}
\label{f:composition}
}
\end{figure*}

The algorithm \CodeIn{NDSarf} below controls Sarf. 
It takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

@@

The arrows in Figure~\ref{f:composition} that transition
from prefix to stem, stem to suffix, and suffix to prefix
require compatibility constraints for 
the transition.

\begin{table}[tb]
\centering
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{tabular}
\label{a:ndsarf}
\end{table}

\section{Islamic Literature Case Study}
\label{sec:islamic}

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on that Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an exhaustive mechanism of authentication
check as. This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

The controller has four states that correspond to 
an abstract position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to the state NAME\_S whenever
NAME is reported by Sarf.
State NRC\_S means that the controller thinks it is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}}
\caption{The controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

@@

The controller reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the current sequence,
and marks the end of the hadith with the beginning of the next sequence. 
It also marks the chain of narrators as the sequence itself.

The definition of input labels such as NAME and IBN depends on the 
morphological analyzer. 
However, our case-based controller approach was able to perform well
under both Sarf and a refined version of Sarf.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, and in particular in the context
of the Arabic language, annotations resulting from 
an ongoing branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case based morphological 
analyzer for Arabic text. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
Each alive machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Another difficulty is that  Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a delimiter such as a space in between. 
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

@@

We find evidence to our hypothesis in~\cite{Maamouri:10} where 
a new corpus was added to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}. 
The addition required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new genre of data challenged the existing data processing and 
annotation guidelines. 
Alternative guidelines came to place where the high level NLP 
task interacted with the morphological analyzer. 
This led to refining SAMA~\cite{Kulick:10}.

The work of Habash~\shortcite{Habash:06} reports that different 
types of morphological analyses behave differently on the same case 
study. 
The analyzers may find different solutions or 
may present the solutions in different orders. 
The presented solutions and the order of presentation may not be 
appropriate for the case study. 
We find strong evidence in our results 
that a case-based morphological analyzer that allows a case-based 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

@@

In this paper, we present Sarf, a {\em novel case-based efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by a case based controller.
%Each alive machine in Sarf represents a valid morphological analysis.
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

Sarf uses 
concatenative analysis with a novel 
{\em recursive} 
refinement. 
Sarf introduces recursive affixes in order to
better retain part of speech information and enhance the 
efficiency of affix matching.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
The authentication of a \noArRL{.hady_t} depends on 
the credibility of each of the narrators as reported in 
separate biography books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in 
Figure~\ref{f:exhadith} are proper names and once
connected toghether they form complex names of narrators. 
For example, narrator $n_1$ has the first name 
\noTrRL{qtybT} and his father name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely.% as long 
%as the text is valid.
We are also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of complex structures 
that are three level deep in hierarchy.
%We also showcase that the results of the accurate high level 
%abstraction
%results of the case-based analysis was not affected by a major
%refinement to the analyzer that increased low level analysis accuracy.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix deterministic tansducers. 
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The case-based controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a case-based analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
\transtrue

%{\bf Sarf example.}

@@

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state.

Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

@@

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
%The relation between morphemes and categories is one 
%to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of $s$ is a 
valid morphological analysis if and only if 
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair.

Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. We write $f: a | a f$. 
%Sarf only keeps atomic affixes and their categories.

\subsection{Affix transducers}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
This transducer is deterministic and thus 
its traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The affix transducers encode $L_{pp}$ and
$L_{xx}$.
An accept state in the transducer corresponds to the last letter 
of an affix in the affix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsection{Stem transducer}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names which we 
obtained from online and biblical sources.

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is deterministic and free of cycles where the 
accept states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsection{Construction of Sarf from transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic construction of Sarf
from the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all deterministic and cycle free as discussed 
is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
The cycles are introduced in the abstract machine
for pedagogical purposes.

\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{Sarf NDFSA.}
\label{f:composition}
}
\end{figure}

The algorithm \CodeIn{NDSarf} 
takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

@@

%The arrows in Figure~\ref{f:composition} that transition
%from prefix to stem, stem to suffix, and suffix to prefix
%require compatibility constraints for 
%the transition.

\begin{table}[tb]
%\centering
%\resizebox{.9\columnwidth}{!}{
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{tabular}
%}
\label{a:ndsarf}
\end{table}

\section{Islamic literature case study}
\label{sec:islamic}

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on that Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an exhaustive mechanism of authentication
checks. This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper, we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

@@

Intuitively, the controller reports a valid chain 
of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of 
the current sequence,
and marks the end of the hadith with the beginning of the 
next sequence. 
It also marks the chain of narrators as the sequence itself.

The controller has four states that correspond to 
a position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to state NAME\_S on
NAME.
State NRC\_S means that the controller is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

@@

\begin{figure}[tb!]
\center{
\resizebox{.8\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}}
\caption{The controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

%The definition of input labels such as NAME and IBN depends on the 
%morphological analyzer. 
%However, our case-based controller approach was able to perform well
%under both Sarf and a refined version of Sarf.
